---
title: "Data624 - Project2"
author: "Amanda Arce, Jatin Jain, Amit Kapoor"
date: "5/2/2021"
output:
  html_document:
    fig_width: 15
    highlight: pygments
    number_sections: no
    theme: flatly
    toc: yes
    toc_float: yes
  pdf_document:
    latex_engine: xelatex
    toc: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.align="center", fig.width = 10)
```





# Overview
ABC Beverage has new regulations in place and the leadership team requires the data scientists team to understand the  manufacturing process, the predictive factors and be able to report to them predictive model of PH. The selection of model depends upon various factors like model accuracy, data relevance, cross validation etc.


# R packages
We will use `r` for data modeling. All packages used for data exploration, visualization, preparation and modeling are listed in Code Appendix.


```{r libraries, include=FALSE, warning=FALSE, message=FALSE}
# Libraries
library(readxl)
library(tidyverse)
library(caret)
library(doParallel)
library(DataExplorer)
library(psych)
library(mice)
library(MASS)
library(caret)
library(AppliedPredictiveModeling)
library(lars)
library(pls) 
library(earth)
library(xgboost)
library(Cubist)
library(randomForest)
library(DT)

set.seed(624)
```



# Data Exploration

We will first get the historical dataset, provided in excel and use it to analyze and eventually predict the PH of beverages.

```{r data}
# download training data from git repo
temp.file <- tempfile(fileext = ".xlsx")
download.file(url="https://github.com/DATA624-PredictiveAnalytics-Project2/Project2/blob/main/StudentData.xlsx?raw=true", 
              destfile = temp.file, 
              mode = "wb", 
              quiet = TRUE)

# read excel for training data
train.df <- read_excel(temp.file, skip=0)

# download testing data from git repo
download.file(url="https://github.com/DATA624-PredictiveAnalytics-Project2/Project2/blob/main/StudentEvaluation.xlsx?raw=true", 
              destfile = temp.file, 
              mode = "wb", 
              quiet = TRUE)

# read excel for testing data
test.df <- read_excel(temp.file, skip=0)


# transform Brand.code to factor
train.df$`Brand Code` = as.factor(train.df$`Brand Code`)
test.df$`Brand Code` = as.factor(test.df$`Brand Code`)
```

## Data summary
There are 31 predictor variables that are numeric and 1 predictor variable `Brand Code` which is factor. The training dataset has 2,571 observations.

```{r, glimpse}
glimpse(train.df)
```


```{r, desc}
describe(train.df) %>% dplyr::select(-vars, -trimmed, -mad, -se)
```

Based of above description, we can see the dataset has missing values so it would need imputation. The predictors `Oxygen Filler`, `MFR`, `Filler Speed` and `Temperature` seems highly skewed and would require transformation. This could be seen in below histogram plots as well. 

## Variables Distribution
Below we have shown the distribution of dataset variables. There are 2 sets of histograms; the one in red is natural distribution and the ones in green are logarithmic disctribution

```{r, hist}
plot_histogram(train.df, geom_histogram_args = list("fill" = "tomato4"))
```

```{r, log10-hist}
# log histograms
plot_histogram(train.df, scale_x = "log10", geom_histogram_args = list("fill" = "springgreen4"))
```


## Missing Data

The summary and following graphs show the missing data in training dataset. The plot below shows more than 8% data is missing for `MFR` variable. Next feature that has missing data is `Filler Speed` which shows more than 2% missing data. The missing data will be handled through imputation.

```{r, colsum}
colSums(is.na(train.df))
```


```{r, pltmiss}
plot_missing(train.df[-1])
```



## Correlation

below plot shows the correlation among numeric variables in the dataset. We can see few variables are highly correlated. We will handle the pairwise predictors that has correlation above 0.90 in data preparation section.

```{r, corr}
forcorr <- train.df[complete.cases(train.df),-1]
corrplot::corrplot(cor(forcorr), method = 'ellipse', type = 'lower')
```

## Outliers
In this section we will check the outliers in the data. An outlier is an observation that lies an unusual distance from other values in a random sample.These outlier could impact predictions so will be handled through imputation 

```{r boxplot}

# boxplot 
par(mfrow = c(3,4))
for(i in colnames(train.df[-1])){
boxplot(train.df[,i], xlab = names(train.df[i]),
  main = names(train.df[i]), col="blue", horizontal = T)
}
```



# Data Preparation

## Handling missing and outliers
The very first in data preparation we will perform is handling missing data and outliers through imputation. We will use mice package to perform imputation here. MICE (Multivariate Imputation via Chained Equations) is one of the commonly used package for this activity. It creates multiple imputations for multivariate missing data. Also we will perform `nearZeroVar` to see if a variable has very little change or variation and not useful for prediction. If we founf any predictor variable satisfying this condition we would remove it.


```{r train-impute}
set.seed(317)

# Training set
train.df.clean <- mice(data.frame(train.df), method = 'rf', m=2, maxit = 2, print=FALSE)
train.df.clean <- complete(train.df.clean)

nzv_preds <- nearZeroVar(train.df.clean)
train.df.clean <- train.df.clean[,-nzv_preds]
```

```{r eval-impute}
set.seed(317)

# Testing set
test.df.clean <- mice(data.frame(test.df), method = 'rf', m=2, maxit = 2, print=FALSE)
test.df.clean <- complete(test.df.clean)
```




## Create Dummy Variables
The variable Brand Code is a categorical variable, having 4 classes (A, B, C, and D). For modeling, we got to convert into set of dummy variables. We will use `dummyVars` function for this purpose that creates a full set of dummy variables.

```{r dv-train}
set.seed(317)
dum.brandcode <- dummyVars(PH ~ Brand.Code, data = train.df.clean)
dum.train.predict <- predict(dum.brandcode, train.df.clean)
train.df.clean <- cbind(dum.train.predict, train.df.clean) %>% dplyr::select(-Brand.Code)
```

```{r dv-eval}
set.seed(317)
dum.brandcode <- dummyVars( ~ Brand.Code, data = test.df.clean)
dum.test.predict <- predict(dum.brandcode, test.df.clean)
test.df.clean <- cbind(dum.test.predict, test.df.clean) %>% dplyr::select(-Brand.Code)
```



## Correlation
Next step is to remove highly correlated predictor variables. we will use the cutoff as 0.90 here.


```{r corr-train}
highCorr <- findCorrelation(cor(train.df.clean), 0.90)
train.df.clean <- train.df.clean[, -highCorr]
```


## Preprocess using transformation

In this step we will use caret `preprocess` method using transformation as `YeoJohnson` which applies Yeo-Johnson transformation, like a BoxCox, but values can be negative as well.

```{r transform-train}
set.seed(317)
preproc_traindf <- preProcess(train.df.clean, method = "YeoJohnson")
train.df.clean <- predict(preproc_traindf, train.df.clean)
```

```{r transform-eval}
set.seed(317)
preproc_testdf <- preProcess(test.df.clean, method = "YeoJohnson")
test.df.clean <- predict(preproc_testdf, test.df.clean)
```




## Training and Test Partition
Finally in this step for data preparation we will partition the training dataset for training and validation using `createDataPartition` method from `caret` package. We will reserve 75% for training and rest 25% for validation purpose.


```{r partition}
set.seed(317)

partition <- createDataPartition(train.df.clean$PH, p=0.75, list = FALSE)

# training/validation partition for independent variables
X.train <- train.df.clean[partition, ] %>% dplyr::select(-PH)
X.test <- train.df.clean[-partition, ] %>% dplyr::select(-PH)

# training/validation partition for dependent variable PH
y.train <- train.df.clean$PH[partition]
y.test <- train.df.clean$PH[-partition]
```





# Build Models



## Linear Regression


### Simple Linear Regression

We will start with Simple Linear Regression model. It will include all the predictor variables in training dataset.

```{r lm}
set.seed(317)

lm_model <- lm(y.train ~ ., data = X.train)
summary(lm_model)
```


We can see that Simple Linear Regression model only covers 38% of variability of data. Next we will check for better models which covers better variability, RMSE and MAE. We can consider this Simple regression as a benchmark model among others we are going to check.


### Partial Least Squares

Partial least squares (PLS) is an alternative to ordinary least squares (OLS) regression. It reduces the predictors to a smaller set of uncorrelated components and then performs least squares regression on these components, instead of on the original data. PLS finds linear combinations of the predictors called components. PLS finds components that attempts to maximally summarize the variation of the predictors while at the same time attempts these components to have maximum correlation with the response.


```{r pls}
set.seed(317)

# tune pls model 
pls_model <- train(x=X.train,
                 y=y.train,
                 method="pls",
                 metric="Rsquared",
                 tuneLength=10, 
                 trControl=trainControl(method = "cv")
                 )

pls_model
```


```{r}
pls_model$bestTune
```



```{r plot-pls}
plot(pls_model)
```


```{r rs-pls}
pls_model$results %>% 
  filter(ncomp == pls_model$bestTune$ncomp) %>% 
  dplyr::select(ncomp,RMSE,Rsquared)
```

```{r rsq-pls}
data.frame(Rsquared=pls_model[["results"]][["Rsquared"]][as.numeric(rownames(pls_model$bestTune))],
           RMSE=pls_model[["results"]][["RMSE"]][as.numeric(rownames(pls_model$bestTune))])
```
Rsquared was used to select the optimal model using the largest value. The final value used for the model was ncomp = 10 which corresponds to best tune model. In this case we see that $R^2$ is 0.33 so only covers 33% variability in data but it produces small RMSE.

## Non Linear Regression

### MARS
MARS creates a piecewise linear model which provides an intuitive stepping block into non-linearity after grasping the concept of multiple linear regression. MARS provided a convenient approach to capture the nonlinear relationships in the data by assessing cutpoints (knots) similar to step functions. The procedure assesses each data point for each predictor as a knot and creates a linear regression model with the candidate features

```{r mars, warning=FALSE, message=FALSE}
set.seed(317)
marsGrid <- expand.grid(.degree=1:2, .nprune=2:30)
mars_model <- train(x=X.train, 
                    y=y.train,
                    method = "earth",
                    tuneGrid = marsGrid,
                    trControl = trainControl(method = "cv"))

```


```{r mars-bt}
# final parameters
mars_model$bestTune
```

```{r plot-mars}
# plot RMSE
plot(mars_model)
```

```{r summ-mars}
summary(mars_model$finalModel)
```

```{r rs-mars}
data.frame(Rsquared=mars_model[["results"]][["Rsquared"]][as.numeric(rownames(mars_model$bestTune))],
           RMSE=mars_model[["results"]][["RMSE"]][as.numeric(rownames(mars_model$bestTune))])
```
RMSE was used to select the optimal model using the smallest value. The final values used for the model were nprune = 12 and degree = 1  which corresponds to best tune model. In this case we see that $R^2$ is 0.33 so only covers 33% variability in data but it also produces small RMSE.



### Support Vector Machines
The objective of the support vector machine algorithm is to find a hyperplane in an N-dimensional space (N being the number of features) that classifies the data points. Hyperplanes are decision boundaries to classify the data points. Data points that falls on either side of the hyperplane can be qualified for different classes. Support vectors are data points that are closer to the hyperplane and effect the position and orientation of the hyperplane. Using these support vectors, we do maximize the margin of the classifier.

```{r svm, warning=FALSE, message=FALSE}
set.seed(317)
svm_model <- train(x=X.train, 
                   y=y.train,
                   method = "svmRadial",
                   tuneLength = 10,
                   trControl = trainControl(method = "cv"))

svm_model
```

```{r summ-svm}
summary(svm_model$finalModel)
```


```{r plot-svm}
# plot RMSE
plot(svm_model)
```

```{r rs-svm}
data.frame(Rsquared=svm_model[["results"]][["Rsquared"]][as.numeric(rownames(svm_model$bestTune))],
           RMSE=svm_model[["results"]][["RMSE"]][as.numeric(rownames(svm_model$bestTune))])
```


RMSE was used to select the optimal model using the smallest value. The final values used for the model were sigma = 0.02339058 and C = 4. We see an improvement here in $R^2$ value which is 0.50 so this model covers 50% variability in the data and RMSE is smallest as well among the models used so far.


## Trees

### Single Tree
Regression trees partition a data set into smaller groups and then fit a simple model for each subgroup. Basic regression trees partition the data into smaller groups that are more homogenous against the response. To achieve outcome consistency, regression trees determine the predictor to split on and value of the split, the depth or complexity of the tree and the prediction equation in the terminal nodes

```{r st, warning=FALSE, message=FALSE}
set.seed(317)

st_model <- train(x=X.train,
                  y=y.train,
                  method = "rpart",
                  tuneLength = 10,
                  trControl = trainControl(method = "cv"))

st_model
```

```{r bt-st}
st_model$bestTune
```



```{r plot-st}
# plot RMSE
plot(st_model)
```


```{r rs-st}
data.frame(Rsquared=st_model[["results"]][["Rsquared"]][as.numeric(rownames(st_model$bestTune))],
           RMSE=st_model[["results"]][["RMSE"]][as.numeric(rownames(st_model$bestTune))])
```


RMSE was used to select the optimal model using the smallest value. The final value used for the model was cp = 0.0120765. We see $R^2$ value is 0.40 so this model covers 40% variability in the data and RMSE as 0.133. The Rsquared value is comparatively low as compared to previous best value.



### Boosted Tree
Boosting algorithms are influenced by learning theory. Boosting algorithm seeks to improve the prediction power by training a sequence of weak models where each of them compensates the weaknesses of its predecessors. The trees in boosting are dependent on past trees, have minimum depth and do not contribute equally to the final model. It requires usto specify a weak model (e.g. regression, shallow decision trees etc) and then improves it.

```{r gbm, warning=FALSE, message=FALSE}
set.seed(317)

# boosting regression trees via stochastic gradient boosting machines

gbmGrid <- expand.grid(interaction.depth = c(5,10), 
                       n.trees = seq(100, 1000, by = 100), 
                       shrinkage = 0.1,
                       n.minobsinnode = c(5,10))

gbm_model <- train(x=X.train,
                   y=y.train,
                   method = "gbm",
                   tuneGrid = gbmGrid, 
                   trControl = trainControl(method = "cv"),
                   verbose = FALSE)

gbm_model
```


```{r bt-gbm}
gbm_model$bestTune
```

```{r plot-gbm}
plot(gbm_model)
```

```{r rs-gbm}
data.frame(Rsquared=gbm_model[["results"]][["Rsquared"]][as.numeric(rownames(gbm_model$bestTune))],
           RMSE=gbm_model[["results"]][["RMSE"]][as.numeric(rownames(gbm_model$bestTune))])
```


Tuning parameter 'shrinkage' was held constant at a value of 0.1. RMSE was used to select the optimal model using the smallest value. The final values used for the model were n.trees = 600, interaction.depth = 10, shrinkage = 0.1 and n.minobsinnode = 5. The $R^2$ and RMSE are 0.54 and 0.11 respectively on training data. This is the best Rsquared so far.


### Random Forest
Random forest consists of a large number of individual decision trees that work as an ensemble. Each model in the ensemble is used to generate a prediction for a new sample and these predictions are then averaged to give the forest’s prediction. Since the algorithm randomly selects predictors at each split, tree correlation gets reduces as compared to bagging. In random forest algorithm, we first select the number of models to build and theen loop through this number and train a tree model. Once done then avearage the predictions to get overall prediction. In random forests, trees are created independently, each tree is created having maximum depth and each tree contributes equally in the final model.

```{r rf, warning=FALSE, message=FALSE}
set.seed(317)

rf_model <- train(x=X.train,
                  y=y.train,
                  method = "rf",
                  tuneLength = 10,
                  trControl = trainControl(method = "cv"))

rf_model
```


```{r bt-rf}
rf_model$bestTune
```

```{r plot-rf}
plot(rf_model)
```

```{r rs-rf}
data.frame(Rsquared=rf_model[["results"]][["Rsquared"]][as.numeric(rownames(rf_model$bestTune))],
           RMSE=rf_model[["results"]][["RMSE"]][as.numeric(rownames(rf_model$bestTune))])
```


RMSE was used to select the optimal model using the smallest value. The final value used for the model was mtry = 16. It has $R^2$ as 0.65 and RMSE as 0.10. Both these values are best among the models used so far. 

Lets see the informative variables found by Random Forest models. we will use `varImp` method to find these variables. 

```{r rf-varImp}
varImp(rf_model)
```

```{r rf-plt-vimp}
plot(varImp(rf_model), top=10, main="Random Forest")
```


From above plot, it is evident `Mnf.Flow` is the most informative variable for PH response variable.


### Cubist
Cubist is a rule-based model. A tree is built where the terminal leaves contain linear regression models. These models are based upon the predictors used in previous splits along with intermediate models. The tree is reduced to a set of rules which initially are paths from the top of the tree to the bottom. Rules are eliminated via pruning or combined and the candidate variables for the models are the predictors that were pruned away. .


```{r cubist, warning=FALSE, message=FALSE}
set.seed(317)

cubist_model <- train(x=X.train,
                      y=y.train,
                      method = "cubist",
                      tuneLength = 10,
                      trControl = trainControl(method = "cv"))

cubist_model
```


```{r bt-cubist}
cubist_model$bestTune
```

```{r plot-cubist}
plot(cubist_model)
```



```{r rs-cubist}
data.frame(Rsquared=cubist_model[["results"]][["Rsquared"]][as.numeric(rownames(cubist_model$bestTune))],
           RMSE=cubist_model[["results"]][["RMSE"]][as.numeric(rownames(cubist_model$bestTune))])
```


RMSE was used to select the optimal model using the smallest value. The best tune  for the cubist model which resulted in the smallest root mean squared error was  with 20 committees. It had RMSE = 0.112, and $R^2$ = 0.578. So far, it covered 57%  of the variability in the data than all other variables and with the low RMSE.


### XGB Tree

```{r xgb, warning=FALSE, message=FALSE}
set.seed(123) 

xgb_trcontrol <- trainControl(
  method = "cv",
  number = 5,  
  allowParallel = TRUE,
  verboseIter = FALSE,
  returnData = FALSE
)

# Setting up a grid search for the best parameters
xgbGrid <- expand.grid(nrounds = c(100,200),  # this is n_estimators above
                       max_depth = c(10, 15, 20, 25),
                       colsample_bytree = seq(0.5, 0.9, length.out = 5),
                       eta = 0.1,
                       gamma=0,
                       min_child_weight = 1,
                       subsample = 1
                      )



xgb_model <- train(x=X.train,
                  y=y.train,
                  method = "xgbTree",
                  trControl = xgb_trcontrol,
                  tuneGrid = xgbGrid)

xgb_model
```


```{r bt-xgb}
xgb_model$bestTune
```

```{r plot-xgb}
plot(xgb_model)
```



```{r rs-xgb}
data.frame(Rsquared=xgb_model[["results"]][["Rsquared"]][as.numeric(rownames(xgb_model$bestTune))],
           RMSE=xgb_model[["results"]][["RMSE"]][as.numeric(rownames(xgb_model$bestTune))])
```


# Select Model

To select the best model for making predictions for evaluation data, we will look at 3 parameters.

* $R^2$, which shows the variance explained by given model.
* RMSE (Root Mean Squared Error), which is the std deviation of the residuals.
* MAE (Mean Absolute Error), which is avg of all absoulte errors. 

Here we will summarize the resamplling to compare the above 3 values among all the models followed by checking the prediction on validation data which we reserved earlier during data partition.


```{r sel-model1}
set.seed(317)
summary(resamples(list(PLS=pls_model, MARS=mars_model, SVM=svm_model, RandFrst=rf_model,  Cubist=cubist_model, SingTree=st_model,Boosting=gbm_model)))
```


```{r bwplot}
bwplot(resamples(list(PLS=pls_model, MARS=mars_model, SVM=svm_model, RandFrst=rf_model,  Cubist=cubist_model, SingTree=st_model, Boosting=gbm_model)), main = "Models Comparison")
```






```{r prediction}
set.seed(317)

pls_pred <- predict(pls_model, newdata = X.test)
mars_pred <- predict(mars_model, newdata = X.test)
svm_pred <- predict(svm_model, newdata = X.test)
rf_pred <- predict(rf_model, newdata = X.test)
cubist_pred <- predict(cubist_model, newdata = X.test)
st_pred<- predict(st_model, newdata = X.test)
gbm_pred <- predict(gbm_model, newdata = X.test)

data.frame(rbind(PLS=postResample(pred=pls_pred,obs = y.test),
                 MARS=postResample(pred=mars_pred,obs = y.test),
                 SVM=postResample(pred=svm_pred,obs = y.test),
                 SingTree=postResample(pred=st_pred,obs = y.test),
                 RandFrst=postResample(pred=rf_pred,obs = y.test),
                 Boosting=postResample(pred=gbm_pred,obs = y.test),
                 Cubist=postResample(pred=cubist_pred,obs = y.test)))
```


We can see here Random Forest performed the best among all the models tried considering the 3 metrics Rsquared, RMSE and MAE, we identified earlier.


# Prediction

Based on the analysis so far, it is confirmed that the Random Forest model is the optimal model. we will now use it tp predict PH values of evaluation dataset and then write it in csv.

```{r pred-PH}
set.seed(317)

# remove PH from evaluation data
test.df.clean <- test.df.clean %>% dplyr::select(-PH)

# predict final PH values
test.df.clean$PH <- predict(rf_model, newdata = test.df.clean)
```

Here are the predicted values of PH for evaluation dataset.

```{r show-ph}
# PH predictions
test.df.clean$PH %>% tibble::enframe(name = NULL) %>% datatable()
```


```{r plt-hist-ph}
plot_histogram(test.df.clean$PH)
```


```{r save-PH}
write.csv(test.df.clean$PH, "StudentEvaluations_PHPredictions.csv")
```


# Conclusion

After extracting the data from the given files, we did first perform data exploration which helped us to find missing data, correlation among the variables and outliers. Next we performed steps for data preparation that included handling missing and outliers through mice, creating dummy vars for a categorical variable Brand Code, remove highly correlated variables, transform data for Normality and finally datata partition of 75% and 25% for training and validation respectively. We then trained various models using linear regression, non linear and Trees model. We finally founf the optimal model as Random Forest for predicting PH values for evaluation data. 

We notice that all the values predicted are than 8. This value translates that the beverage made is alkaline. At the start of this study, we were not known about the nature of the ABC Beverage company i.e. what type of beverage manufacturer it was. But from this study we can conclude that this company mainly produces alkaline beverages like water, tea, fruit drinks and all.

# References

* Applied Predictive Modeling. Max Kuhn and Kjell Johnson
* https://machinelearningmastery.com/pre-process-your-dataset-in-r/
* https://www.analyticsvidhya.com/blog/2016/03/tutorial-powerful-packages-imputing-missing-values/
* https://newalbanysmiles.com/ph-values-of-common-beverages/


# Code Appendix

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```





